### 2012-11-09 inst/examples/comparing-methods.Rmd
updated examples using sequential approach

```{r}
X <- seq(-5,5,len=50)
obs <- data.frame(x = c(-4, -3, -1,  0,  2),
                  y = c(-2,  0,  1,  2, -1))
l <- 1
sigma_n <- 0.8
```

@@

Direct method 
```{r}
  cov_xx_inv <- solve(K + sigma_n^2*I)
  Ef <- cov(X, obs$x) %*% cov_xx_inv %*% obs$y
  Cf <- cov(X, X) - cov(X, obs$x)  %*% cov_xx_inv %*% cov(obs$x, X)
```

```{r}
gp <- gausspr(obs$x, obs$y, kernel="rbfdot", kpar=list(sigma=1/(2*l^2)), fit=FALSE, scaled=FALSE, var=0.8)
y_p <- predict(gp, X)
```

Things that should be equivelent but aren't quite:

```{r}
ggplot(data.frame(x=X, Ef=Ef, Y=Y, y_p = y_p))+ geom_point(aes(x,Ef), col='red') + geom_point(aes(x,Y)) + geom_line(aes(x,y_p))
````

>>>>

```{r}
X <- seq(-5,5,len=50)
obs <- data.frame(x = c(-4, -3, -1,  0,  2, 3),
                  y = c(-2,  0,  1,  2, -1, -1))
#x <- -5:5
#obs <- data.frame(x = -5:5, y = sin(x) + rnorm(length(x),sd=.1))
l <- 1
sigma_n <- 0.8
```

@@

Direct method

``` {r}
  cov_xx_inv <- solve(K + sigma_n^2*I)
  Ef <- cov(X, obs$x) %*% cov_xx_inv %*% obs$y
  Cf <- cov(X, X) - cov(X, obs$x)  %*% cov_xx_inv %*% cov(obs$x, X)
```

  Direct sequential method, avoids matrix inverse instability
``` {r}
ef <- numeric(length(X))
cf <- matrix(0, nrow=length(X), ncol=length(X))
for(i in 1:length(obs$x)){
  ef <- ef + cov(X, obs$x[i]) * obs$y[i] / as.numeric(cov(obs$x[i],obs$x[i]) + sigma_n^2)
  cf <- cf + cov(X, X) - cov(X, obs$x[i])  %*% cov(obs$x[i], X) / as.numeric(cov(obs$x[i],obs$x[i]) + sigma_n^2)
}
```

``` {r}
ggplot(data.frame(x=X, Ef=Ef, ef=ef))+ geom_point(aes(x,Ef), col='red') + geom_line(aes(x,ef))
````

kernlab method

``` {r}
gp <- gausspr(obs$x, obs$y, kernel="rbfdot", kpar=list(sigma=1/(2*l^2)), fit=FALSE, scaled=FALSE, var=0.8)
y_p <- predict(gp, X)
```

mlegp method

``` {r}
require(mlegp)
out <- mlegp(obs$x, obs$y, nugget = 1)
y_m <- predict(out, as.matrix(X))
```

Compare these results:

``` {r}
require(reshape2)
df <- data.frame(x = X, Ef = Ef, Y = Y, y_p = y_p, ef = ef, y_m = y_m)
df <- melt(df, id = "x")
ggplot(df)+ geom_jitter(aes(x, value, color = variable)) + geom_point(data = obs, aes(x,y))
````

Compare alpha from Cholesky and alpha from `kernlab` to $K(x,x)^{-1} \vec y$

### 2012-11-09 inst/examples/gaussian-process-basics.Rmd
updated examples using sequential approach

$$x|y \sim \mathcal{N}\left(E,C)$$
$$E = cov(X_p, X_o) cov(X_o,X_o)^{-1} y$$
$$C= cov(X_p, X_p) - cov(X_p, X_o) cov(X_o,X_o)^{-1} cov(X_o, X_p) \right)$$

@@

```{r}
sigma.n <- 0.25
cov_xx_inv <- solve(cov(obs$x, obs$x) + sigma.n^2 * diag(1, length(obs$x)))
Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% obs$y
Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
```

@@

fig2c <- ggplot(dat,aes(x=x,y=value)) +
  geom_ribbon(data=NULL, 
              aes(x=x_predict, y=Ef, ymin=(Ef-2*sqrt(diag(Cf))), ymax=(Ef+2*sqrt(diag(Cf)))),
              fill="grey80") + # Var
  geom_line(aes(color=variable)) + #REPLICATES
  geom_line(data=NULL,aes(x=x_predict,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
fig2c
```

@@

>>>>

$$x|y \sim \mathcal{N}(E,C)$$
$$E = cov(X_p, X_o) cov(X_o,X_o)^{-1} y$$
$$C= cov(X_p, X_p) - cov(X_p, X_o) cov(X_o,X_o)^{-1} cov(X_o, X_p)$$

@@

```{r}
sigma.n <- 0.3
cov_xx_inv <- solve(cov(obs$x, obs$x) + sigma.n^2 * diag(1, length(obs$x)))
Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% obs$y
Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
```

@@

```{r}
dat <- data.frame(x=x_predict, t(values))
dat <- melt(dat, id="x")
Ey <- data.frame(x=x_predict, y=(Ef), ymin=(Ef-2*sqrt(diag(Cf))), ymax=(Ef+2*sqrt(diag(Cf))))

fig2c <- ggplot(dat,aes(x=x,y=value)) +
  geom_ribbon(data=Ey, aes(x=x,y=y, ymin=ymin, ymax=ymax), fill="grey80") + # Var
  geom_line(aes(color=variable)) + #REPLICATES
  geom_line(data=Ey,aes(x=x,y=y), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x") 
fig2c + opts(panel.background = theme_rect(fill = "transparent"), 
        plot.background = theme_rect(fill = "transparent"))
```

@@

<!-- Continuing... 2012-10-22 -->

The likelihood is given by 
$$\log(p(y | X)) = -\tfrac{1}{2} \mathbf{y}^T (K + \sigma_n^2 \mathbf{I})^{-1} y  - \tfrac{1}{2} \log\left| K + \sigma_n^2 \mathbf{I} \right| - \tfrac{n}{2}\log 2 \pi$$

Which is:

```{r}

#' observation pairs, obs$x, obs$x, 
#' SE covariance between two observations 
minusloglik <- function(pars){

  l <- pars[1]
  sigma.n <- pars[2]

  cov <- function(X, Y) outer(X, Y, SE, l)
  I <- diag(1, length(obs$x))
  K <- cov(obs$x, obs$x)

  0.5 * t(obs$y) %*% solve(K + sigma.n^2 * I) %*% obs$y +
    log(det(K + sigma.n^2*I)) +
    length(obs$y) * log(2 * pi) / 2
  }
```

```{r}
pars <- c(l=1, sigma.n=1)
o <- optim(pars, minusloglik)
o$par
```

## A comparison to built-in R utilities

Some basic Gaussian process tools are provided in the `kernlab` package,

```{r}
library(kernlab)
```

through the `gausspr` function,

```{r}
gp <- gausspr(obs$x, obs$y, kernel="rbfdot", kpar=list(sigma=1/(2*l^2)), fit=FALSE, scaled=FALSE, var=0.8)
```

Where `rbfdot` indicates the Gaussian (Radial basis function) kernel we use above, with parameter `sigma=1/(2l^2)` (note that gausspr defines the kernel parameter differently, see:

```{r}
gp@kernelf@.Data
```
note this takes vectors `x`, `y` and returns a scalar, not the covaraiance matrix).  The function will try to estimate hyperparameters such as `sigma` if they are not given unless `fit=FALSE`.  The data is also scaled to 0 mean, unit variance unless `scaled=FALSE` is set.

We can then use the estimated process to infer values on a specified grid through `predict`, e.g. for the example above,

```{r}
y_p <- predict(gp, x_predict)
dat2 <- data.frame(x = x_predict, variable = "Y_p", value = y_p)
```

```{r}
ggplot(dat2, aes(x, value)) + geom_point() + geom_line(data=Ey,aes(x=x,y=y), size=1) 
```

### Fitting hyperparameters by maximum likelihood

If we do not specify `fit=false` then `gausspr` will estimate the hyperparamters by maximum likelihood:

```{r}
gp_fit <- gausspr(obs$x, obs$y, kernel="rbfdot", kpar=list(sigma=1/(2*l^2)), scaled=FALSE, var=0.8)
```
